---
title: "ACA Market Share Modeling"
author: "Aaron Olcerst"
date: "2022-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
require(readxl)
library(ggplot2)
library(lme4)
library(knitr)
library(lattice)
defaultW <- getOption("warn")
options(warn = -1)
```

## Prep market enrollment

Here we're collecting data describing the number of people enrolled across all carriers in each county for each year.  This will serve as the denominator for our market share calculations

```{r}
source("enrollment_munging.R")
mkt <- enrollment_munging()
head(mkt)

```

## Prep carrier-level enrollment

This section collects carrier-specific enrollment for each county from 2014-2020.  This will serve as our numerator for the market share calculations.

```{r}
source("signup_munging.R")
signups <- signup_munging()
head(signups)
```

## Prep price position data

Next we need pricing data.  Because the enrollment data is so course, we'll need to aggregate to relative price positions for each carrier's metal tier at the county level for each year in the data set.

```{r}
source("price_munging.R")
prices <- price_munging()
head(prices)
```

## Combine data sets and prep for modeling

We're almost there!  Just need to combine all the data we've created so far.

```{r}
source("data_prep.R")
data <- data_prep()
head(data)
```
## EDA

Now that we have the data prepped, let's look at some visuals. Checking out data from Ohio, we can see that market shares tend to cluster along different silver price position slopes for each carrier.

```{r}
source("visualization.R")
plot_OH_state()
```

In Florida, it's easy to see that the Blues can sustain higher membership with a less competitive price position
```{r}
plot_FL_state()
```
## Build a model

This next step builds a logistic GLM using silver price position, bronze price position, and carrier x state dummy variables

```{r}
source("model_build.R")
glm.mod <- build_model()
```
## What can we learn from the model?

We can now build illustrative plots that show, when controlling for price position, the implicit differences in the competitiveness of each carrier

```{r}
viz_dat <- prep_lines()
plot_illustrative()
```

However, it's worth noting that this model isn't great in its current state. If we look at county-level plots of a single carrier, Ambetter, it's obvious that there is a lot of county-level variation that this model can't account for.

```{r}
plot_OH_amb()

```

```{r}
plot_OH_amb()

```
## Final thoughts

The best way to deal with this is to build a model that incorporates county-level variation without over-fitting.  This could be through a non-linear mixed-effect model, or a carefully constructed tree-based model.

Ultimately, however, much better models can be constructed with first-party data, which not only allows for metal-tier-level projections, but also delineation into separate models which project retention of existing members and acquisition of new members.

